---
title: "03: Datasets"
execute:
  error: true
embed-resources: true
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
rstudioapi::viewer('https://r-training.netlify.app/tutorials/psychrlogy/01_fundRmentals/03_datasets')
```

### Open in a New Browser Tab

```{r}
utils::browseURL('https://r-training.netlify.app/tutorials/psychrlogy/01_fundRmentals/03_datasets')
```



## Overview


## Setup


### Projects


### Documents


#### Quarto documents


#### Scripts


### Installing and Loading Packages

Load the {tidyverse} package in your Quarto document.

```{r}

```



## Reading In


### Reading from File

Run the `here::here()` function to see what it does.
```{r}

```



Use `here::here()` to generate a file path to the `syn_data` data file.

Then, use `readr::read_csv()` to read in the `syn_data.csv` file and store the result in an object called `syn_data`.

```{r}

```



#### Reading from URL


Read the CSV file hosted at <https://raw.githubusercontent.com/drmankin/practicum/master/data/syn_data.csv> and save it to the object name `syn_data`.

```{r}

```



## Codebook


## Viewing


### Call the Object


Call the `syn_data` object to see what it contains.

```{r}

```



### A Glimpse of the Data


Use `dplyr::glimpse()` to get a glimpse of your dataset.

```{r}

```



### View Mode

Open the `syn_data` dataset using the `View()` function **in the Console**.

```{r}

```



Using **only** View mode, figure out the following:

1. What is the range of the variable `gc_score`?
1. How can you arrange the dataset by score in `scsq_imagery`?
1. How many participants had "Yes" in the variable `syn_graph_col`?
1. Which `gender` category had more participants?
1. Of the participants who said "Yes" to `syn_seq_space`, what was the highest SCSQ technical-spatial score?

```{r}

```



## Overall Summaries


### Basic Summary


Print out a summary of `syn_data` using the `summary()` function.

```{r}

```



### Other Summaries


Print out a summary of `syn_data` using the `datawizard::describe_distribution()` function.

```{r}

```



**CHALLENGE**: There are some variables missing from this output. What are they? Why aren't they included?

```{r}

```



## The Pipe


## Describing Datasets


Using the native pipe, print out the number of columns and the names of those columns in the `syn_data` dataset.

*Hint*: This will be two separate commands!

```{r}

```



Using the native pipe, save the number of participants in the `syn_data` dataset in a new object of your choice.

```{r}

```



## Describing Variables


### Counting


Using the `syn_data` dataset, produce a tibble of counts of how many participants had any kind of synaesthesia. Then, produce a second tibble, adding in gender as well.

*Hint*: Use the [codebook](03_datasets.qmd#codebook) to find the variables to use.

```{r}

```



### Subsetting


Subset `syn_data` using `$` to get out all the values stored in the `scsq_organise` variable.

```{r}

```



Subset `syn_data` using `dplyr::pull()` to get out all the values stored in the `gc_score` variable. How would you read this code?

```{r}

```



### Descriptives


Calculate the mean, standard deviation, and median of the SCSQ global subscale, and the range of the grapheme-colour synaesthesia score.

Try using each subsetting method at least once.

```{r}

```



### Visualisations


Try making a histogram and a boxplot, using any of the variables in the `syn_data` dataset. Try using `$` and `pull()` once each.

*Optionally*, if you feel so inclined, use the help documentation to spruce up your plots a bit, such as changing the title and axis labels.

```{r}

```



**CHALLENGE**: Try making a barplot and a scatterplot.

For the barplot, make a visualisation of how many people are synaesthetes or not (regardless of synaesthesia type).

For the scatterplot, choose any two SCSQ measures.

Both of these require some creative problem-solving using the help documentation and the skills and functions covered in this tutorial.

```{r}

```


