---
title: "01/02: IntRoduction"
format: html
editor: visual
execute:
  warning: false
  message: false
  error: true
embed-resources: true
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
rstudioapi::viewer('https://r-training.netlify.app/tutorials/psychrlogy/01_fundRmentals/01_02_intro')
```

### Open in a New Browser Tab

```{r}
utils::browseURL('https://r-training.netlify.app/tutorials/psychrlogy/01_fundRmentals/01_02_intro')
```



## Orientation


### The RStudio Interface


#### Source


#### Console


## Errors

Type literally any gibberish, words, keysmashes etc. into the code chunk in the workbook and press Run (or Ctrl/Cmd + Shift + Enter).

```{r}

```



### GlossoRlia: the Language of Errors


## Types of Data


### Numeric Data

Type any single number and run the code.

```{r}

```


How does R handle commas within a number (e.g. to separate the thousands place from the hundreds)? How about full stops for decimals?

```{r}

```


Add together your shoe size and the number of windows in the room you're currently in.

```{r}

```


Try subtracting, dividing, and multiplying the same two numbers.

```{r}

```



#### Vectors {#vectors}

Print out every whole number between 1 and 50.

```{r}

```


Print out all the numbers 12 through 30; all of the numbers 23 through 55; and 36, all in one command.
```{r}

```



#### Vector Calculations

Create a vector of every whole number between 37 and 63, and subtract 7 from each element.

```{r}

```



### Character Data

Create a vector containing the first five animals you think of, then print the 3rd one.

```{r}

```



### Logical Data

Write the following assertions in R:

-   5 is greater than 10
-   6 is less than 12
-   27 is less than or equal to 27
-   49 does not equal 93
-   420 equals 42

```{r}

```


Use a single command to ask R whether the numbers 2 through 10 are less than or equal to 6.

```{r}

```



## Class and Coercion

Use the `class()` function to get R to print the values `"numeric"`, `"logical"`, and `"character"`.

```{r}

```


What data type does R give you if you combine numbers and characters in `c()`?

```{r}

```


Use an `as.*()` function to convert the following vector of participant ages into numeric data: `c(20, "42", "36 years old")`. What do you think will happen to each element?

```{r}

```



## Objects


### Creating an Object


#### Naming Objects

Think of a research scenario familiar to you with two independent groups. You're welcome to draw from your own research or expertise, but you should choose something with numerical scores. Some ideas include:

1.  Reaction times on a button-pressing task from a control and an experimental group
2.  Statistics anxiety scores from first and second year UG students
3.  Quiz marks from students with practicals scheduled 9am and students with practicals at 6pm

Make a note of the scenario you chose. Then, create two new objects: one that contains a vector of six scores from the first of the two groups, and the second that has six different scores from the second group.

*Hint*: Just make up some numbers that sound plausible!

```{r}

```



### Calling an Object

Call both of the objects you just created.

```{r}

```



### Using Objects

Calculate the mean of each of the two sets of scores you created.

```{r}

```


Calculate the **difference** in the mean of each of the two sets of scores, and save this difference in a new object called `quiz_diff`.

```{r}

```


What is the class of these objects?

```{r}

```



### Overwriting Objects

First, let's imagine we get three new participants in each condition of our previous study. Update the same two objects you created previously with three new scores each.

```{r}

```



## Functions


### Basics and Help

Try running the `round()` function.

```{r}

```


Open the help documentation for the `round()` function by running `?round()` or `help(round)` in the Console.
```{r}

```



### Arguments


#### Unnamed Arguments


#### Named Arguments


### Using Functions

Use the `round()` function to round 64.3333333 to two decimal places.

If you prefer, you can do this with one of the means you calculated for your own scores earlier.

```{r}

```



#### Passing Multiple Values to Arguments

Before you go on, have a go using a single `round()` command to round 64.3333333, 59.5452, and 0.198 at once.

*Hint*: Refer to [Vectors](#vectors).
```{r}

```



## Quick Test: *t*-test

Bring up the help documentation for `t.test()` and use it to run a *t*-test comparing your two sets of scores.

```{r}

```


Using the help documentation, re-run the *t*-test with equal variances assumed.

```{r}

```


