---
title: "02: IntRoduction II"
execute:
  error: true
embed-resources: true
editor: visual
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
#| eval: false
rstudioapi::viewer('https://r-training.netlify.app/tutorials/docs/02_intro')
```

### Open in a New Browser Tab

```{r}
#| eval: false
utils::browseURL('https://r-training.netlify.app/tutorials/docs/02_intro')
```

## Overview

## Setup

### Projects {#projects}

### Documents

#### Quarto documents

Click the "Render" button at the top of your worksheet Quarto document to see what it produces.

```{r}

```

#### Scripts

## Packages {#packages}

### Loading `tidyverse`

Load the {tidyverse} package in your workbook.

*Note*: If you are on the Cloud workspace, {tidyverse} will already be installed. If not, you may need to install it first.

```{r}

```

## Functions

### Basics and Help

As a warm-up, calculate the mean of the four numbers given above and store this number in a new object called `mean_to_round`.

```{r}

```

Try running the `round()` function.

```{r}

```

Open the help documentation for the `round()` function by running `?round()` or `help(round)` **in the Console**.

```{r}

```

### Arguments

### Using Functions

Use the `round()` function to round 23.15575 (stored in `mean_to_round`) to two decimal places.

```{r}

```

### Passing Multiple Values to Arguments

Before you go on, have a go using **a single `round()` command** to round 23.15575, 59.5452, and 0.198.

*Hint*: Refer to [Vectors](#vectors).

```{r}

```

## Packages Revisited

Round the number .00793 to three decimal places using the `rd()` function.

```{r}

```

### Installing Packages

Install the {weights} package. Replace `function_name` in the example command below and run this command **in the Console** only.

*Note*: As mentioned above, please don't regularly install packages if you're on the Posit Cloud workspace - but in this specific case, it's good to get the practice!

\[Use the Console for this task!\]

Round the number .00793 to three decimal places using the `rd()` function.

```{r}

```

### Loading Packages

Round the number .00793 to three decimal places using the `rd()` function, using either method to access the function in the {weights} package.

```{r}

```

## Exercises

Run the following code in your workbook to generate the data you need for the following tasks.

```{r}
practice_data <- c(15, 784, 2, NA, 956, 9, 23, 8, 326, 1, 406)
```

Return the median and range of `practice_data`.

*Hint*: If you are getting back a result that isn't particularly informative, use the help documentation to figure out how to deal with it.

```{r}

```

Calculate a 10% trimmed mean for `practice_data`, but specify the trim first in the function.

```{r}

```

Go back to the code you wrote for the exercises above and add some comments to your code in plain English.

```{r}

```
