---
title: "05: Filter and Select"
embed-resources: true
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
rstudioapi::viewer('https://r-training.netlify.app/tutorials/psychrlogy/02_essentials/05_filter')
```

### Open in a New Browser Tab

```{r}
utils::browseURL('https://r-training.netlify.app/tutorials/psychrlogy/02_essentials/05_filter')
```



## Overview


## Setup


### Packages

Load the necessary packages.

```{r}

```



### Data

Read in the dataset and save it in a new object, `anx_data`.

On the Cloud, you can read in this dataset from the `data` folder using `here::here()`.

Elsewhere, you can download the dataset, or copy the dataset URL, from the [Data and Workbooks page](../../../data_workbooks.qmd).

```{r}

```



#### Codebook


## Filter


### General Format


### Filtering with Assertions

Produce a subset of `anx_data` that doesn't contain any male participants.

```{r}

```


Produce a subset of `anx_data` that contains only participants as old as the median age, or younger.

```{r}

```



### Multiple Assertions

Filter `anx_data` to keep only cases where the value of `rmars_s_test2` is between 2 and 4.

*Hint*: You can use two separate assertions to do this, or check out `dplyr::between()`.

```{r}

```



### Data Cleaning


#### Pre-Exclusions


Remove any preview runs from the dataset, keeping only real data.

```{r}

```



#### Recording Exclusions

Follow along with the following data cleaning steps, trying them out in a code chunk for yourself as you go. You'll need them at the end!
```{r}

```



#### Consent


#### Age


#### Missing Values


##### Single Variable


##### All Variables


**CHALLENGE**: Why has every single row in the dataset been dropped? Using any method you like, investigate what's happened.

```{r}

```



#### Reporting

**CHALLENGE**: Using the objects counting intial, final, and excluded cases and what we covered last time about inline code, write a brief journal-style description of your exclusion process.

What is the benefit of taking the extra effort to store these counts in objects? Under what circumstances might this be (particularly) useful?

```{r}

```



## Select


### General Format


### Selecting Directly

Create a subset of `anx_data` that contains the following variables:

-   The participant's age
-   The first variable in the original dataset
-   All of the STARS variables

```{r}

```


Create a subset of `anx_data` that contains the following variables:

-   All of the original variables but NOT `distribution`
-   `mcq` renamed `condition`

```{r}

```



### Using {tidyselect}

Open the help documentation by running `?dplyr::select` in the Console to see examples of how to use all of the `<tidyselect>` helper functions.
```{r}

```



Select the variables in `anx_data` that have to do with state anxiety.

```{r}

```


Select all the variables in `anx_data` that are NOT the R-MARS, R-MARS-S, or STICSA.

```{r}

```



**CHALLENGE**: Select all the `stars` variables but NOT the `stars_m` variables.

```{r}

```



#### Using Functions


#### Using Custom Functions

**CHALLENGE:** Select the variables in `anx_data` that are character type, or that do NOT contain any missing values.

*Hint*: You may need to use function(s) that we haven't covered in the tutorials so far to solve this.

```{r}

```



## Quick Test: *t*-test redux

Bring up the help documentation for `t.test()` and use it to run a *t*-test comparing responses for the first item of the STICSA post-test state anxiety scale between people who saw the maths MCQs vs the stats MCQs.

*Hint*: Try using the formula notation, just like we did in the previous tutorial for `lm()`.

```{r}

```


**CHALLENGE**: Using the help documentation, re-run the *t*-test with equal variances assumed.

```{r}

```


**CHALLENGE**: Using what you known about `papaja`, inline code, and Quarto, report the results of this *t*-test in full without typing any of the numerical results into the text directly.

Then, render your workbook to see the results!

```{r}

```


