---
title: "07: Visualisations"
embed-resources: true
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
rstudioapi::viewer('https://r-training.netlify.app/tutorials/psychrlogy/02_essentials/07_dataviz')
```

### Open in a New Browser Tab

```{r}
utils::browseURL('https://r-training.netlify.app/tutorials/psychrlogy/02_essentials/07_dataviz')
```



## Overview


### What do UGs know?


## Setup


### Packages

Load the packages.

```{r}

```



### Data

Read in the dataset and save it in a new object, `anx_score_data`.

On the Cloud, you can read in this dataset from the `data` folder using `here::here()`.

Elsewhere, you can download the dataset, or copy the dataset URL, from the [Data and Workbooks page](../../../data_workbooks.qmd).

```{r}

```



#### Codebook


## Grammar of Graphics


### Layers


### Mapping


### Geoms, etc.


## Histograms and Density Plots

Follow along with the instructions below to build your own plot as we go. Feel free to tweak details like colour and labels as you like.
```{r}

```



### Adjusting Binwidth


### Colour and Fill Manual


### Adjusting Axes with `scale_*()`


### Adding a Theme

Create a density plot of the `rmars_s_test_score` variable (or another continuous variable in the dataset of your choosing). For a finished plot, make sure you:

-   Choose colours for your plot.
-   Adjust the labels, breaks, and limits as necessary.
-   Add a theme.

```{r}

```



### Facet Wrap


## Barplots

Create a new variable in the dataset, `stars_help_cat`, that contains either "low" for people who scored below 3 on the STARS Asking for Help subscale, or "high" for people who scored 3 or above. Make sure to save your changes back to the dataset.

```{r}

```


Follow along with the instructions to build your own plot as we go. Feel free to tweak details like colour and labels as you like.
```{r}

```



### Position Dodge


### Reordering Categories


#### Option 1: Scale Labels


#### Option 2: Factor Levels

Finish up your plot by adjusting the following elements:

-   Give a name to each axis
-   Relabel discrete categories where appropriate
-   Adjust continuous limits and breaks where appropriate
-   Choose new colours!

*Hint*: You'll have three axes to adjust. Use the help documentation for each if you get stuck.

```{r}

```



## Quick Test: $\chi^2$


Using the `chisq.test()` help documentation, perform a $\chi^2$ test of association for the `stars_help_cat` and `gender` variables.

```{r}

```



**CHALLENGE**: Use a matrix to run this test, instead of two individual vectors.

*Hint*: The way I've solved it, this requires reading the help documentation carefully and using a new function to create a `table()`.

```{r}

```



**CHALLENGE**: Report the results of the $\chi^2$ test in full, including a check of whether the expected frequencies are all greater than 5.

```{r}

```



## Raincloud and Violin Plots

Follow along with the instructions below to build your own plot as we go. Feel free to tweak details like colour and labels as you like.
```{r}

```



### Calculating Stats

In second year, students are also introduced to violin plots. Adapt the code we've already written for raincloud plots to create a violin plot with means and CIs instead.

```{r}

```


**CHALLENGE**: The raincloud plot we produced above was serviceable but not very *aesthetic*. See if you can adapt that code to reproduce the plot below.

*Hints*: Run `vignette("ggrain")` in the Console for a friendly tour of the options for raincloud plots. The colours are all from the Sussex colour palette.


```{r}

```



## Scatterplots

Follow along with the instructions below to build your own plot as we go. Feel free to tweak details like colour and labels as you like.
```{r}

```



### Line of Best Fit


### Colour and Fill Palette

**CHALLENGE**: In the code above, I had to repeat a lot of elements for both `fill` and `colour`. See if you can figure out how to remove the need to change the name and labels within the plotting code by making a change to the dataset beforehand.

*Hint*: You might try a non-standard variable name with backticks!

```{r}

```



### 3D Plots


## Reporting with Quarto


### Captions and Alt Text


### Cross-Referencing

