---
title: "09: Labelled Data and Qualtrics"
embed-resources: true
editor: visual
---

## Open the Tutorial

Use the following code chunks to open the accompanying tutorial.

### Open in RStudio Viewer

```{r}
#| eval: false
rstudioapi::viewer('https://r-training.netlify.app/tutorials/docs/09_qtrics')
```

### Open in a New Browser Tab

```{r}
#| eval: false
utils::browseURL('https://r-training.netlify.app/tutorials/docs/09_qtrics')
```



## Overview


### Acknowledgements


## Qualtrics


### SAV Data


### Setting Up Qualtrics


#### Blocks


#### Questions


##### Questions in R


##### Matrix Questions


### Library


### Exporting Data


## Setup


### Packages

Load the packages.

```{r}

```



### Data

Read in the `mil_data.sav` object from folder, or alternatively from Github via URL, as you prefer.

On the Cloud, you can read in this dataset from the `data` folder using `here::here()`.

Elsewhere, you can download the dataset, or copy the dataset URL, from the [Data and Workbooks page](../../data_workbooks.qmd).

```{r}

```



### Codebook


## Variable Names


### Default Variable Names


### Renaming 


#### Option 1: Rename in Qualtrics


#### Option 2: `rename()`


#### Option 3: `rename_with()`


## Exercises: Names


Clean up your dataset by doing the following. You can do the steps in whatever order works for you.

- Keep all the date variables, demographic questions, and items measuring Global Meaning, Mattering, and Belonging.
- Rename any default-named Qualtrics variables (starting with "Q") to a sensible name.

Refer to the [Codebook](#codebook) to figure out which variables are which.

```{r}

```



## Labelled Data


### Working with Labels


### Variable Labels


#### Getting Labels


#### Creating/Updating Labels


#### Searching Labels


### Value Labels


#### Getting Labels


#### Creating/Updating Labels


### Missing Values


## Exercises: Labels


Identify the item that mentions 'job'. Then, change the variable label of this item so that it just says 'Occupational Status'

```{r}

```



For the `income` variable, change the value label 'I prefer not to disclose information about my annual income as part of this research study.' to 'Prefer not to say'

```{r}

```



**CHALLENGE**: Above we converted the value label for the value 3 in the `demo_gender` variable to "Other" by assigning a new label wholesale. How can you do this with a regular expression instead?

If you did already update the value label using the code above, reset it to its original state by running the code below first.


```{r}

```



## Data Dictionaries


## Converting Variables


### Factors


#### Revision of Factors


#### Converting to Factors


### Numeric


### Efficient Conversion


## Exercise: Conversion and Wrangling


Prepare the `mil_data` dataset for analysis.

1. Produce a final data dictionary and save it.
2. Convert all categorical variables to factor, and all scale rating variables to numeric.
3. Reverse-code `global_meaning_2`.
4. Create composite scores for all of the subscale variables.

As a reminder, we covered reverse-coding and composite scores in the tutorial on `mutate()`.

```{r}

```



## Well done!

